Galois Slicing as Differentiation

### 2025-07-11 main/introduction.tex
[ tex ] write up the semantics and some examples
[ tex ] fix bib
Fix borked line break.

  \GPS associates with each run of the program a Galois connection telling us how the inputs and outputs are related in that run. The backwards portion $\partial (\mathit{query}\,l)_r$ tells us, given an approximation of the output, what the least approximation of the input is needed to generate that output. In the case of the two runs considered in this example, if we say we are not interested in the output by feeding in the least approximation $\bot$, then we find that we only need the least approximation of the input:
  \begin{displaymath}
    \partial (\mathit{query}\,l)_r(\bot) = [(\mathsf{a},\bot), (\mathsf{b}, \bot), (\mathsf{a}, \bot)]
  \end{displaymath}
  for both $l = \mathsf{a}$ and $l = \mathsf{b}$. If instead we take the greatest approximation of the output (i.e., the output ``$1$'' itself), then the two query runs' backwards approximations return different results:
  \begin{displaymath}
    \begin{array}{l}
      \partial (\mathit{query}\,\mathsf{a})_r(1) = [(\mathsf{a},0), (\mathsf{b},\bot), (\mathsf{a},1)] \\
      \partial (\mathit{query}\,\mathsf{b})_r(1) = [(\mathsf{a},\bot), (\mathsf{b},1), (\mathsf{a},\bot)]
    \end{array}
  \end{displaymath}
  Pieces of the input that were {\em not} used are replaced by $\bot$. As we expect, the run of the query with label $\mathsf{a}$ depends on the entries in the database labelled with $\mathsf{a}$, and likewise for the run with label $\mathsf{b}$.

  In this case, the forwards portion of the Galois connection tells us for each approximation of the input whether or not it is sufficient to compute the output. If we provide insufficient data to compute the output, then we will get an underapproximated output. For example, $\partial (\mathit{query}\, \mathsf{a})_f([(\mathsf{a},0),(\mathsf{b},\bot),(\mathsf{a},\bot)]) = \bot$ because we need all the values associated with the label $\mathsf{a}$ to compute their sum.

@@

\bob{Have stolen this text for \autoref{sec:diff-stab-summary}}
\begin{itemize}
\item For \GPS, we assume that every value has an associated lattice of {\em approximations}. For differentiable programs, every point has an associated vector space of {\em tangents}.
\item For \GPS, every program has an associated forward approximation map that takes approximations forward from the input to the output. This map {\em preserves meets}. For differentiable programs, every program has a forward derivative that takes tangents of the input to tangents of the output. The forward derivative map is {\em linear}, so it preserves addition of tangents and the zero tangent.
\item For \GPS, every program has an associated backward approximation map that takes approximations of the output back to least approximations of the input. This map {\em preserves joins}. For differentiable programs, every program has a reverse derivative that takes tangents of the output to tangents of the input. This map is again {\em linear}.
\item For \GPS, the forward and backward approximation maps are related by being a Galois connection. For differentiable programming, the forward and reverse derivatives are related by being each others' transpose.
\end{itemize}

@@

>>>>

  \GPS associates with each run of the program a Galois connection telling us how the inputs and outputs are related in that run. The backwards portion $\partial (\mathit{query}\,l)_r$ tells us, given an approximation of the output, what the least approximation of the input is needed to generate that output. In the case of the two runs considered in this example, if we say we are not interested in the output by feeding in the least approximation $\bot$, then we find that we only need the least approximation of the input:
  \begin{displaymath}
    \partial (\mathit{query}\,l\,\mathit{db})_r(\bot) = [(\mathsf{a},\bot), (\mathsf{b}, \bot), (\mathsf{a}, \bot)]
  \end{displaymath}
  for both $l = \mathsf{a}$ and $l = \mathsf{b}$. If instead we take the greatest approximation of the output (i.e., the output ``$1$'' itself), then the two query runs' backwards approximations return different results:
  \begin{displaymath}
    \begin{array}{l}
      \partial (\mathit{query}\,\mathsf{a}\,\mathit{db})_r(1) = [(\mathsf{a},0), (\mathsf{b},\bot), (\mathsf{a},1)] \\
      \partial (\mathit{query}\,\mathsf{b}\,\mathit{db})_r(1) = [(\mathsf{a},\bot), (\mathsf{b},1), (\mathsf{a},\bot)]
    \end{array}
  \end{displaymath}
  Pieces of the input that were {\em not} used are replaced by $\bot$. As we expect, the run of the query with label $\mathsf{a}$ depends on the entries in the database labelled with $\mathsf{a}$, and likewise for the run with label $\mathsf{b}$.

  In this case, the forwards portion of the Galois connection tells us, for each approximation of the input, whether or not it is sufficient to compute the output. If we provide insufficient data to compute the output, then we will get an underapproximated output. Here for example we will find that $\partial (\mathit{query}\, \mathsf{a})_f([(\mathsf{a},0),(\mathsf{b},\bot),(\mathsf{a},\bot)]) = \bot$ because we need all the values associated with the label $\mathsf{a}$ to compute their sum.

@@

\begin{itemize}
\item For \GPS, we assume that every value has an associated lattice of {\em approximations}. For differentiable programs, every point has an associated vector space of {\em tangents}.
\item For \GPS, every program has an associated forward approximation map that takes approximations forward from the input to the output. This map {\em preserves meets}. For differentiable programs, every program has a forward derivative that takes tangents of the input to tangents of the output. The forward derivative map is {\em linear}, so it preserves addition of tangents and the zero tangent.
\item For \GPS, every program has an associated backward approximation map that takes approximations of the output back to least approximations of the input. This map {\em preserves joins}. For differentiable programs, every program has a reverse derivative that takes tangents of the output to tangents of the input. This map is again {\em linear}.
\item For \GPS, the forward and backward approximation maps are related by being a Galois connection. For differentiable programming, the forward and reverse derivatives are related by being each others' transpose.
\end{itemize}

@@

\todo{Rewrite this to match the actual contents}
